---
title: "Census Income Exploratory Data Analysis"
author: "Ben Stano"
date: "12/14/21"
output: 
  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "",
  fig.pos = "",
  message = FALSE,
  tidy = "styler",
  warning = FALSE,
  fig.align = "center",
  highlight = TRUE
)

library(tidyverse)
library(broom)
library(GGally)
library(corrr)
library(car)
```

# Description of Each Variable

## Response
* median_income: the median income of all people within a tract

## Predictors
The predictors in the census data were given as a flat number of individuals within the tract who meet the given criteria. We transformed the variables into proportions of a tracts total population to normalize the data and make it more suitable for regression.

* white: percentage of white people in the given tract 
* bachelors: percentage of people with a bachelor's degree in the given tract
* graduate: percentage of people with a graduate degree in the given tract
* same_residence: percentage of people who have been at the same residence for the past year in a given tract
* private_transport: percentage of people who ride in private transport to work in the given tract
* own_car: percentage of people who drive their own car to work in the given tract
* public_transport: percentage of people who take public transit to work in the given tract
* bike: percentage of people who ride a bike to work in the given tract
* walk: percentage of people who walk to work in the given tract
* native: percentage of people who are US natives in the given tract
* foreign_noncitizen: foreigner living in the US who is not a US citizen
* foreign_naturalized: foreigner living in the US who is a US citizen
* SSI: Social Security Income, percentage of people who use SSI

## Unused Variables
* GEOID: Unique GEOID of the given tract
* state: the state which the given tract is in
* tract: the tract that the row's data pertains to
* county: the county the given tract is in
* population: the population of a given tract. While this was not used directly as a predictor, it was used to normalize the predictors.



```{r}
raw_data <- readRDS("census_raw_area.RDS")
glimpse(raw_data)
```

# Exploratory Data Analysis

All variables have some extreme outliers. Skew is apparent even after scaling and limiting the range of the plot.
```{r}
census_n <- select(raw_data, where(is.numeric))
census_n |>
  map_dfc(scale) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") |>
  ggplot(aes(x = Value, color = Variable)) +
  geom_density(show.legend = FALSE) +
  geom_line(aes(x = Inf, y = Inf)) +
  lims(x = c(-2, 5)) +
  guides(legend.title.position = "top") +
  theme(legend.position = "bottom") +
  labs(x = "Standard Deviations", y = "Density")
```

Median income has a somewhat positive-skewed distribution, as income is always greater than or equal to zero, though this violates no assumptions.
Taking state medians of median income, it is closer to normal.
```{r}
census_n |>
  ggplot(aes(median_income)) +
  geom_density()

group_by(raw_data, state) |>
  summarize(median_income = median(median_income, na.rm = TRUE)) |>
  ggplot(aes(x = 0, y = median_income, label = state)) +
  geom_violin(fill = "skyblue") +
  geom_point()
```

A pairs plot reveals the skew of many of the distributions, even after normalizing for population and removing unpopulated tracts. While some relationships are strong, they don't appear linear.

Normalizing Data by Population
```{r}
census_normalized <- filter(census_n, population > 0, !is.na(median_income)) %>%
  mutate(across(where(is.numeric) & -c(population, median_income), ~ .x / population)) %>%
  filter(area != 0)
```

Inspecting correlation
```{r}
# setting up the variables that will be used for linear modeling
# since population is of census tracts, which are not uniform in size, it would be difficult to find a meaningful interpretation of population
census_lm <- census_normalized %>%
  select(-population)

corry <- correlate(census_lm)
corry

stretch(corry) %>%
  filter(abs(r)>0.5)
```
A correlation plot. The strongest correlations with median income are 
among the education-related variables, which is unsurprising.

```{r}
expand_grid(row = colnames(census_lm), col = row) |>
  mutate(Correlation = c(cor(census_lm, use = "complete.obs"))) |>
  ggplot(aes(x = fct_rev(row), y = col, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(colors = c("red", "white", "blue"), breaks = c(-1, -.5, 0, .5, 1)) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "Correlation Plot", x = "", y = "")
```


# Modeling

I have to drop one variable that created a linear dependency, as well
as some metadata.
```{r}
# fitting the initial linear model
initial <- lm(median_income ~ ., data = census_lm)
tidied <- tidy(initial)
tidied
```

The curvature and non-constant variance of the plot clearly suggests a transformation is needed for the response variable.

```{r}
aout <- augment(initial)
ggplot(aout, aes(.fitted, .resid))+ 
  geom_hline(yintercept=0) +
  geom_point(alpha = .05)

# adding log of median income as a column in census_lm
census_lm <- census_lm %>%
  mutate(log_median_income = log(median_income))
# fitting a new linear model with the log of median income
logged_y <- lm(log_median_income ~ . - median_income -SSI, data=census_lm)
tidy(logged_y)
```

Logging the response leaves a roughly linear residual plot, though with some inconsistency in variance. The QQ-Plot inevitably shows serious issues related to the zero bound for median income, but outside of this the plot is approximately linear.
```{r}
aout_log <- augment(logged_y)
ggplot(aout_log, aes(.fitted, .resid))+ 
  geom_hline(yintercept=0) +
  geom_point(alpha = .05)
```


```{r}
ggplot(aout_log, aes(sample = `log_median_income`)) +
  geom_qq() +
  geom_qq_line()
```
Taking a look at the SSE reduction from each variable
```{r}
Anova(logged_y)
```
# Transforming Area

Area does not seem to be performing well, could a transformation make it more significant?

Logging area
```{r}
census_lm
census_lm <- mutate(census_lm, log_area = log(area))

logged_y <- lm(log_median_income ~ . - median_income -SSI -area, data=census_lm)
Anova(logged_y)
```

The log of area performs much better than area by itself. Let's remove variables that don't seem to be contributing

# Attempting to Create a Person of Color Variable 

Seeing if a merged predictor of black population percentage and native american merged would preform.

```{r}
logged_y_revised <- lm(log_median_income ~ black + native_american + hispanic_latino + bachelors + graduate + same_residence + bike + walk + foreign_naturalized + log_area, data = census_lm)
Anova(logged_y_revised)
```


```{r}
census_lm <- mutate(census_lm, poc = black + native_american)

logged_y_POC <- lm(log_median_income ~ poc + hispanic_latino + bachelors + graduate + same_residence + bike + walk + foreign_naturalized + log_area, data = census_lm)
Anova(logged_y_POC)
```




Refitting the model without parameters which did not show consequential impact within the parameters of our model.
```{r}
logged_y_revised <- lm(log_median_income ~ poc + bachelors + graduate + same_residence + private_transport + bike + walk + log_area, data = census_lm)
Anova(logged_y_revised)
```

```{r}
logged_y_revised <- lm(log_median_income ~ poc + bachelors + graduate + same_residence + private_transport + log_area + bike + walk + foreign_naturalized, data = census_lm)
Anova(logged_y_revised)
```
```{r}
logged_y_revised <- lm(log_median_income ~ black + bachelors + graduate + same_residence + private_transport + bike + walk + log_area, data = census_lm)
Anova(logged_y_revised)
```

```{r}
logged_y_revised <- lm(log_median_income ~ poc + bachelors + graduate + same_residence + private_transport + log_area + bike + walk + foreign_naturalized, data = census_lm)
Anova(logged_y_revised)
```

Checking the revised parameter residuals. With the removal of numerous predictor variables, the regression model is largely unchanged.
```{r}
aout_logr <- augment(logged_y_revised)
ggplot(aout_logr, aes(.fitted, .resid)) + 
  geom_hline(yintercept=0) +
  geom_point(alpha = .05)
```

Revised parameter qqplot, The results are not meaningfully altered by the removal of excess variables.
```{r}
ggplot(aout_logr, aes(sample = `log_median_income`)) +
  geom_qq() +
  geom_qq_line()
```
Checking the revised parameter model. The comparatively low impact of being white is surprising, perhaps we should see if the result will hold true for other racial parameters.
```{r}
Anova(logged_y_revised)
```
Mapping a pair plot with our preliminary parameters, obviously bachelors and graduate are heavily correlated as a graduate degree has a dependency on first obtaining a bachelors, so some categorical manipulation could be worthwhile, perhaps splitting the two variables into "bachelors and not graduate" and graduate.
```{r}
census_lm |>
  select(log_median_income, black, bachelors, graduate, private_transport, same_residence, bike, walk) |>
  ggpairs(mapping = aes(alpha = .03))
```
# Attempting further predictor transformations

There is a slight increase in SSR for log(SSI+1) over SSI, but not enough to justify the lose of model interpretability that such a transformation would cause.

## Exploring bike transformations
```{r}

# there are many tracts with 0 bike usage
census_lm %>%
  filter(bike==0) %>%
  nrow()

qplot(x=log(bike), y=log_median_income, data=census_lm)

census_lm_trans <- census_lm %>%
  mutate(log_bike = log(bike+1))

logged_y_revised_trans <- lm(log_median_income ~ black + bachelors + graduate + same_residence + private_transport + log_bike + walk, data = census_lm_trans)

Anova(logged_y_revised_trans)
```
There is a slight increase in SSR for log(bike+1) over bike, but not enough to justify the lose of model interpretability that such a transformation would cause.

**There is a large increase in SSR** for log(foreign_naturalized+.001) over foreign_naturalized. There may be enough increase in performance to justify the transformation

## Exploring graduate transformations
```{r}
# there are many tracts with 0 graduate students usage
census_lm %>%
  filter(graduate==0) %>%
  nrow()
qplot(x=graduate, y=log_median_income, data=census_lm)
qplot(x=log(graduate), y=log_median_income, data=census_lm)

census_lm_trans <- census_lm %>%
  mutate(log_graduate = log(graduate+1))

logged_y_revised_trans <- lm(log_median_income ~ black + bachelors + log_graduate + same_residence + private_transport + bike + walk, data = census_lm_trans)

Anova(logged_y_revised_trans)
```
There is a slight increase in SSR for log(graduate+1) over graduate, but not enough to justify the lose of model interpretability that such a transformation would cause.


There is a slight increase in SSR for log(bachelors+1) over bachelors, but not enough to justify the lose of model interpretability that such a transformation would cause.


# Inspecting the Model

```{r}
tidy(logged_y_revised)
```
$$ Log(Income_i) = 8.6 - 0.20 * Black + 1.5*Bachelors + 2.2 *Graduate  $$


```{r}
aout <- augment(logged_y_revised)
qplot(x=.fitted, y=.resid, data=aout) + 
  geom_hline(yintercept=0)
```

# Validating the Model

# Conclusions

















